Best Practices for Alchemical Free Energy Calculations [Article v\versionnumber]

### 2021-03-16 paper/manuscript.tex
Last fixes.

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this propose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

@@

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
\end{itemize}

@@

In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
\begin{equation}\label{eq:delta-delta-G-physical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
\end{split}
\end{equation}
Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
\begin{equation}\label{eq:delta-delta-G-alchemical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
\end{split}
\end{equation}
where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
Note that, when $A$ and $B$ have different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.

@@

In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).

We can rigorously account for fluctuations in other thermodynamic parameters such as changes in volume $V$ when simulating at constant pressure $p$ or changes in number of molecules $N_i$ of species $i$ at constant chemical potential $\mu_i$ (e.g., number of waters or ions) by introducing the \textit{reduced potential}~\cite{shirts2008statisticallya}
\begin{equation}\label{eq:reduced-potential}
u(\vec{q};\vec{\lambda}) \equiv \beta \left[ U(\vec{q};\vec{\lambda}) + p \, V(\vec{q}) + \sum_i \mu_i \, N_i(\vec{q}) + \cdots \right] \, .
\end{equation}
Here, the collection of thermodynamic and alchemical parameters $\{\beta, \vec{\lambda}, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
The free energy of mutating $A$ to $B$ in any environment (e.g., binding site, solvent) can then be computed as
\begin{equation}\label{eq:delta-G-env}
    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, .
\end{equation}
While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of configurations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.

@@

Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest, by introducing multiple intermediate alchemical states at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a interpolation of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1})
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates. This is achieved by using the configurations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
We first note that the accuracy of any free energy calculation method will be dependent on the quality of the underlying force field.  Therefore, any description of the force field for the molecules under study must be carefully checked to be sufficiently accurate to experiment. In particular, one must make sure that if using automatically generated molecular descriptions from either one's own workflow or some other computational chemistry program, there are not obvious problems in these files either through errors in the workflow, or lack of chemical coverage in the data used to construct the molecular description. Torsional parameters, in particular can be misassigned or improperly parameterized.

Given reasonably accurate, current generation force fields correctly implemented, alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

@@

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead in some cases to wide difference in free energies.  Even within this restriction,
finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e.with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol, using different initial velocities. The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat run as well as different velocities.

@@

In lead optimization guided by free energy calculations, we will likely only
synthesize and experimentally test molecules that are predicted to
have favorable free energy changes. We are thus interested in how
often that a molecule predicted to bind stronger actually turns out to
bind stronger. In other words, we are interested in the conditional
probability:
\begin{equation}
  P(\Delta\Delta G<0|\Delta\Delta G^\dagger<0).
  \label{eqn:true-positive}
\end{equation}

@@

Available computational resources and timescales of motion also factor
into this initial analysis. An individual free energy calculation
involves simulations at many different intermediate states (perhaps
20-40 or more) and each of these must typically be long enough to
capture the relevant motions in the system. If such motions are
microsecond events or longer, the computational cost of running 20-40
microsecond or longer simulations for each of $N$ ligands will likely be
prohibitive for most users with today's hardware. Alternatively, if key motions are fast and minimal (as is often assumed in practice), only much shorter simulations may be necessary.

@@

\subsection{Capturing experimental conditions}
\label{subsec:exp_condition}
The calculations aim to capture the alchemical change from one ligand to another as accurately as possible. Therefore, it is necessary to consider details of the experimental setup, such as pH. Biological assays are usually run at neutral pH but this is not always the case. For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other from neutral pH. Therefore, computational protein and ligand preparation protocols should reflect experimental pH.

@@

In summary, the successful use of alchemical calculations not only in, but particularly for drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved. Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study. 
 There are many accounts of success of alchemical calculations, the methods show good performance towards the goal of binding free energy prediction. However, it is important to have realistic expectations.

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding free energy predictions can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models~\cite{cappel2016relative} and GPCRs~\cite{deflorian2020accurate,lenselink2016predicting} for instance, as well as enabling charge change and scaffold hopping~\cite{chen2018accurate, wang2017accurate}, but these systems are undoubtedly more difficult. In the meantime, the use cases are expanding to resistance prediction, selectivity prediction , solubility prediction – an exciting future for alchemical calculations~\cite{hauser2018predicting, albanese2020structure, mondal2019free}.

@@

\begin{itemize}
    \item Do I have one or multiple good receptor structures? (e.g. a good resolution X-ray crystal of the protein target)
    \item Do I have information on one or all of the ligand binding sites? (e.g. a X-ray structure)
    \item Should I include buried waters, or other small molecules that can be found in an X-ray structure?
    \item Are my ligands part of a congeneric series? (i.e. simple R group substitutions around the same scaffold)
    \end{itemize}

@@

\paragraph{Protonation states depend on the pH of the experimental assay}
Care should be taken when preparing ligands and proteins to match the pH of the experimental assay, if known. As mentioned above in Sec.~\ref{subsec:exp_condition}, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands. Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improvements, 2020schrodinger}. Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionized, but care is still needed as the local environment can modify expected ionization states, such as  the catalytic Asp dyad in proteases. Histidine is notoriously more difficult to predict as its pKa suggests it ionizes closer to the experimental pH range. For ligands often the pKa needs to be determined, if not known experimentally. There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}. Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools~\cite{isik2018pka}. While often it can be assumed that the protonation state of a ligand and protein will remain the same as a ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}. BACE, for example, famously undergoes a protonation state change on ligand binding.

@@

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}). In cases where the end states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations. In situations where the end state topologies differ significantly, or there is relatively little spatial overlap between the two end states some user intervention may be necessary to produce a satisfactory input topology.

>>>>

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

@@

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
\end{itemize}

@@

In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
\begin{equation}\label{eq:delta-delta-G-physical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
\end{split}
\end{equation}
Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
\begin{equation}\label{eq:delta-delta-G-alchemical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
\end{split}
\end{equation}
where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
Note that, when $A$ and $B$ have a different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.

@@

In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occurs at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).

We can rigorously account for fluctuations in other thermodynamic parameters such as changes in volume $V$ when simulating at constant pressure $p$ or changes in number of molecules $N_i$ of species $i$ at constant chemical potential $\mu_i$ (e.g., number of waters or ions) by introducing the \textit{reduced potential}~\cite{shirts2008statisticallya}
\begin{equation}\label{eq:reduced-potential}
u(\vec{q};\vec{\lambda}) \equiv \beta \left[ U(\vec{q};\vec{\lambda}) + p \, V(\vec{q}) + \sum_i \mu_i \, N_i(\vec{q}) + \cdots \right] \, .
\end{equation}
Here, the collection of thermodynamic and alchemical parameters $\{\beta, \vec{\lambda}, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
The free energy of mutating $A$ to $B$ in any environment ($\Delta G_{\mathrm{env}}$ e.g., binding site, solvent) can then be computed as
\begin{equation}\label{eq:delta-G-env}
    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, ,
\end{equation}
over the configurational space of the environment ($\Gamma_{\mathrm{env}}$).
While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of configurations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.

@@

Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest by introducing multiple intermediate alchemical states at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a interpolation of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}),
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators, such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya}, can exploit similarities between states to improve the precision of the estimates. This is achieved by using the configurations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
We first note that the accuracy of any free energy calculation method will depend on the quality of the underlying force field.  Therefore, any description of the force field for the molecules under study must be carefully checked to be sufficiently accurate to experiment. In particular, one must make sure that if using automatically generated molecular descriptions from either one's own workflow or some other computational chemistry program, there are no obvious problems in these files either through errors in the workflow, or lack of chemical coverage in the data used to construct the molecular description. Torsional parameters, in particular can be misassigned or improperly parameterized.

Alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

@@

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.  Even within this restriction,
finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e.with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol, using different initial velocities. The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat run as well as different velocities.

@@

In lead optimization guided by free energy calculations, we will likely only
synthesize and experimentally test molecules that are predicted to
have favorable free energy changes. We are thus interested in how
often a molecule predicted to bind stronger actually turns out to
bind stronger. In other words, we are interested in the conditional
probability:
\begin{equation}
  P(\Delta\Delta G<0|\Delta\Delta G^\dagger<0).
  \label{eqn:true-positive}
\end{equation}

@@

Available computational resources and timescales of motion also factor
into this initial analysis. An individual free energy calculation
involves simulations at many different intermediate states (perhaps
20-40 or more) and each of these must typically be long enough to
capture the relevant motions in the system. If such motions are
microsecond events or longer, the computational cost of running 20-40
microsecond or longer simulations for each of $N$ ligands will likely be
prohibitive for most users with today's hardware. On the other hand, if key motions are fast and minimal (as is often assumed in practice), much shorter simulations may be sufficient.

@@

\subsection{Capturing experimental conditions}
\label{subsec:exp_condition}
The calculations aim to capture the alchemical change from one ligand to another as accurately as possible. Therefore, it is necessary to consider details of the experimental setup, such as pH. Biological assays are usually run at neutral pH but this is not always the case. For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other than neutral pH. Therefore, computational protein and ligand preparation protocols should reflect experimental pH.

@@

In summary, the successful use of alchemical calculations, particularly for drug discovery, requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved. Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study. 
 There are many accounts of success of alchemical calculations, the methods show good performance towards the goal of binding free energy prediction. However, it is important to have realistic expectations.

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding free energy predictions can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models~\cite{cappel2016relative} and GPCRs~\cite{deflorian2020accurate,lenselink2016predicting} for instance, as well as enabling charge change and scaffold hopping~\cite{chen2018accurate, wang2017accurate}, but these systems are undoubtedly more difficult. In the meantime, use cases are expanding to resistance prediction, selectivity prediction , solubility prediction – an exciting future for alchemical calculations~\cite{hauser2018predicting, albanese2020structure, mondal2019free}.

@@

\begin{itemize}
    \item Do I have one or multiple good receptor structures? (e.g. a good resolution X-ray crystal of the protein target)
    \item Do I have information on one or all of the ligand binding sites? (e.g. an X-ray structure)
    \item Should I include buried waters, or other small molecules that can be found in an X-ray structure?
    \item Are my ligands part of a congeneric series? (i.e. simple R group substitutions around the same scaffold)
    \end{itemize}

@@

\paragraph{Protonation states depend on the pH of the experimental assay}
Care should be taken when preparing ligands and proteins to match the pH of the experimental assay, if known. As mentioned above in Sec.~\ref{subsec:exp_condition}, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands. Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improvements, 2020schrodinger}. Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably be predicted to be ionized, but care is still needed as the local environment can modify expected ionization states, such as  the catalytic Asp dyad in proteases. Histidine is notoriously more difficult to predict as its pKa suggests it ionizes closer to the experimental pH range. For ligands, often the pKa needs to be determined, if it is not known experimentally. There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}. Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools~\cite{isik2018pka}. While often it can be assumed that the protonation state of a ligand and protein will remain the same as the ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}. BACE~\cite{kim2015conformational}, for example, famously undergoes a protonation state change on ligand binding.

@@

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}). In cases where the end states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations. In situations where the end state topologies differ significantly, or where there is relatively little spatial overlap between the two end states, some user intervention may be necessary to produce a satisfactory input topology.
